---
title: "STAT 1261/2260 Homework 3"
author: "Jason Lee"
date: "2022-09-30"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Needed Libraries Loaded**
```{r}
library(ggplot2)
library(mdsr)
library(babynames)
library(Hmisc)
library(tidyverse)
```

## Lab Recreation
 
### Q1
```{r}
BabynamesDist <- make_babynames_dist()

joseph <- BabynamesDist %>% filter(name == "Joseph", sex == "M")
name_plot <- ggplot(data = joseph, aes(x = year))
name_plot <- name_plot +
  geom_bar(aes(y = count_thousands * alive_prob),
           stat = "identity", 
           fill = "#b2d7e9", colour = "white")
name_plot <- name_plot + geom_line(aes(y = count_thousands), size = 2) 
name_plot <- name_plot +
  ylab("Number of People (thousands)") + xlab(NULL)

median_yob <-
  with(joseph, wtd.quantile(year, est_alive_today, probs = 0.5))

name_plot <- name_plot +
  geom_bar(stat = "identity", colour = "white", fill = "#008fd5",
           aes(y = ifelse(year == median_yob, est_alive_today / 1000, 0)))
name_plot +
  ggtitle("Age Distribution of American Boys Named Joseph") +
  geom_text(x = 1935, y = 40, label = "Number of Josephs \n born each year") +
  geom_text(x = 1915, y = 13, label =
    "Number of Josephs \n born each year \n estimated to be alive \n on 1/1/2014",
    colour = "#b2d7e9") +
  geom_text(x = 2003, y = 40,
    label = "The median \n living Joseph \n is 37 years old",
            colour = "darkgray") +
  geom_curve(x = 1995, xend = 1974, y = 40, yend = 24,
    arrow = arrow(length = unit(0.3,"cm")), curvature = 0.5) + ylim(0, 42)
```

### Q2
```{r}
BabynamesDist <- make_babynames_dist()

lillian <- BabynamesDist %>% filter(name == "Lillian", sex == "F")
name_plot <- ggplot(data = lillian, aes(x = year))
name_plot <- name_plot +
  geom_bar(aes(y = count_thousands * alive_prob),
           stat = "identity", 
           fill = "#b2d7e9", colour = "white")
name_plot <- name_plot + geom_line(aes(y = count_thousands), size = 2) 
name_plot <- name_plot +
  ylab("Number of People (thousands)") + xlab(NULL)

median_yob <-
  with(lillian, wtd.quantile(year, est_alive_today, probs = 0.5))

medText <- paste('The median living Lillian \n is ', max(lillian$year)-median_yob, ' years old')

name_plot <- name_plot +
  geom_bar(stat = "identity", colour = "white", fill = "#008fd5",
           aes(y = ifelse(year == median_yob, est_alive_today / 1000, 0)))
name_plot +
  ggtitle("Age Distribution of American Girls Named Lillian") +
  geom_text(x = 1920, y = 15, label = "Number of Lillians \n born each year") +
  geom_text(x = 1995, y = 12, label =
    "Number of Lillians \n born each year \n estimated to be alive \n on 1/1/2014",
    colour = "#b2d7e9") +
  geom_text(x = 1950, y = 23,
    label = medText,
            colour = "darkgray") +
  geom_curve(x = 1950, xend = 1966, y = 20, yend = 2,
    arrow = arrow(length = unit(0.3,"cm")), curvature = 0.2) + ylim(0, 30)

```


## Boston Housing Data

```{r}
library(MASS)
data("Boston")
```
### Q3
**Dataset Head:**
```{r}
head(Boston)
```
**Dataset Dimensions:**
```{r}
dims <- dim(Boston)
paste('Rows: ', dims[1], ' x ', 'Cols: ', dims[2])
```
**Dataset Names:**
```{r}
names(Boston)
```

### Q4
```{r}
bostonMedv <- ggplot(Boston, aes(x=medv, y = ..density..))+
  geom_histogram(color = 'blue', fill = 'blue')+
  geom_density(color = 'red')

bostonMedv + labs(title = "Histogram of Median House Values in Boston", x = "Median House Value (in $1000)")
```

### Q5
```{r}
bostonRm <- ggplot(Boston, aes(x=rm, y = ..density..))+
  geom_histogram(color = 'black', fill = 'white')+
  geom_density(color = 'purple')

bostonRm + labs(title = "Histogram of Avg Number of rooms per Dwelling\nin Boston", x = "Avg Number of Rooms")
```

### Q6
```{r}
scatterP <- ggplot(Boston, aes(x = medv, y = rm))+
  geom_point()+
  geom_smooth()

scatterP + labs(title = 'Median House Value vs Avg number of Rooms\nper House in Boston', x = 'Median House Value ($1000s)', y = 'Avg Number Rooms')
```

### Q7
```{r}
rm_cat = as.factor(round(Boston$rm))
Boston2 <- data.frame(Boston$medv, rm_cat)
head(Boston2)
```
Boxplot:
```{r}
ggplot(Boston2, aes(y = Boston.medv, x = rm_cat))+
  geom_boxplot()
```

#### **Why do we need to Convert?**
If we don't convert rounded rm values into a categorical variable with the factor() function, we will have a continuous range of rm values that would each be considered its own group. The resulting boxplot graph doesn't show side-by-side comparisons in this case, and only shows a *single* boxplot ranging over the entire x domain (or all the rm values). 
