---
title: 'STAT1261/2260 HW4'
author: 'Jason Lee'
date: '10-4-2022'
output: word_document
---
```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```
##Load the necessary Libraries
```{r}
library(tidyverse)
library(Hmisc)
library(mdsr)
library(babynames)

BabynamesDist <- make_babynames_dist()
paste("Libraries Loaded!")
```

## P1: Median Ages for 25 Most Common Female Names
```{r}
com_fem <- BabynamesDist %>% #com_fem stands for "common female"
filter(sex == "F") %>%
group_by(name) %>%
filter(est_alive_today>0) %>%
summarise(
est_num_alive = sum(est_alive_today),
q1_age = wtd.quantile(age_today, est_alive_today, probs = 0.25),
median_age = wtd.quantile(age_today, est_alive_today, probs = 0.5),
q3_age = wtd.quantile(age_today, est_alive_today, probs = 0.75)
) %>%
arrange(desc(est_num_alive)) %>%
head(25)
com_fem
```
#### Q1
For all living Margarets, the 25th percentile of age is **49**, the median age is **64**, and the 75th percentile age is **75** according to the graphic. These statistics are outdated though as margaret is no longer among the 25 most common female names in the dataset.

#### Q2
The name "Elizabeth" has the largest IQR (75th% - 25th%).

#### Q3
They are ordered in ascending order of median age of females with the 25 most common names.

#### Q4
1) Length: The length of the bars for each name represent the amount of distribution of ages for that particular name.
2) Position: The location of each bar relative to the others suggests similarity or difference in age between names.
3) Direction: Ordering median age in descending order creates a low age -> high age trend in the graph.
4) Color: Using a red dot to represent the median on top of a lighter yellow color for the bar makes it easy to identify the location.

