---
title: "STAT 1261/2260 Homework 8"
author: "Jason Lee"
date: "2022-11-11"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Prep
```{r}
#install.packages(c("NHANES", "tidyverse", "glmnet"))
library(NHANES)
library(tidyverse)
library(glmnet)
library(rpart)
library(rpart.plot)

?NHANES
```

```{r}
data1<- NHANES %>%
  select(HardDrugs,Gender,HealthGen,Depressed,SleepHrsNight,AlcoholDay,SmokeNow,Marijuana) %>% 
  na.omit()
glimpse(data1)
```

#### Problem 1
```{r}
contTable <- with(data1, table(Marijuana, HardDrugs))
propcontTable <- contTable / sum(contTable)
round(propcontTable,3)
```

## Regularized Logistic Regression 1
```{r}
set.seed(99)
train1 <- data1 %>% sample_frac(size = 0.8)
test1 <- data1 %>% setdiff(train1)
```
```{r}
library(glmnet)
form<- as.formula(
  "HardDrugs ~Gender+HealthGen+Depressed+SleepHrsNight+AlcoholDay+SmokeNow+Marijuana")
predictors <- model.matrix(form, data = train1) 
cv.fit <- cv.glmnet(predictors, train1$HardDrugs, family = "binomial", type = "class")
cv.fit$lambda.1se

plot(cv.fit)
```
```{r}
lambda_opt=cv.fit$lambda.1se
mod_lr <- glmnet(predictors, train1$HardDrugs, family = "binomial", lambda = lambda_opt)

form2 <- as.formula(HardDrugs~Gender+Depressed+AlcoholDay+Marijuana)
mod_lr2 <- glm(form2, data=train1,family=binomial)
summary(mod_lr2)
```
#### Problem 2
"MarijuanaYes" or Marijuana is the most significant as it has a near zero p-value (2e-16 ~ 0).

#### Problem 3
There is a strong, positive correlation between Marijuana and HardDrugs as seen by the estimated 3.11 coefficient of Marijuana (HardDrugs = 3.11 * Marijuana + error).

#### Problem 4
Males are more likely to use HardDrugs seen by the moderate, positive estimated coefficient for "GenderMale".

## Decision Tree
```{r}
mod_tree <- rpart(form2,data=train1)
rpart.plot(mod_tree)
```

#### Problem 5
Marijuana usage ("Marijuana = No") is the most important predictor as the root of the decision tree. It has the greatest impact on predicting response. 

#### Problem 6
Those who have tried Marijuana (**Not** Marijuana = No) and are Depressed (**Not** Depressed = None) are most likely to use hard drugs.

## Data Prep
```{r}
data2<- NHANES %>%
  select(Gender,Age,Race1,Education,MaritalStatus,HHIncomeMid,Poverty,HomeOwn,Weight,Height,
         BMI,Pulse,BPSysAve,BPDiaAve,Diabetes,HealthGen,DaysPhysHlthBad,DaysMentHlthBad,
         Depressed,SleepHrsNight,SleepTrouble,AlcoholDay,Smoke100,Marijuana,HardDrugs) %>% 
  drop_na()

train2 <- data2 %>% sample_frac(size = 0.8, fac=HardDrugs)
test2 <- data2 %>% setdiff(train2)
```

## Regularized Logistic Regression 2
```{r}
library(glmnet)
form_full<- as.formula(HardDrugs~Gender+Age+Race1+Education+MaritalStatus+HHIncomeMid+Poverty+HomeOwn+Weight+Height+BMI+Pulse+BPSysAve+BPDiaAve+Diabetes+HealthGen+DaysPhysHlthBad+DaysMentHlthBad+Depressed+SleepHrsNight+SleepTrouble+AlcoholDay+Smoke100+Marijuana)
predictors <- model.matrix(form_full, data = train2) 
cv.fit <- cv.glmnet(predictors, train2$HardDrugs, family = "binomial", type = "class")
cv.fit$lambda.1se

plot(cv.fit)
```
```{r}
lambda_opt=cv.fit$lambda.1se
mod_lr2 <- glmnet(predictors, train2$HardDrugs, family = "binomial", lambda = lambda_opt)
mod_lr2$beta
```

#### Problem 7
```{r}
mod_tree2 <- rpart(form_full, data = data2)
rpart.plot(mod_tree2)
```

*The Decision Tree identified 14 important predictors.*

## Model Assessment
```{r echo = FALSE}
confusion_matrix <- function(data,y,mod){
  confusion_matrix <- data %>% 
  mutate(pred = predict(mod, newdata = data, type = "class"),
         y=y) %>%
  select(y,pred) %>% table()
}
misclass <- function(confusion){
  misclass <- 1- sum(diag(confusion))/sum(confusion)
  return(misclass)
}

logistic.misclassrate <- function(dataset, y, fit, form){
  misclass_lr <- dataset %>% 
  mutate(pred.logistic = predict(fit, newx = model.matrix(form, data = dataset), 
         type = "class")) %>% 
  mutate(misclassify = ifelse(y != pred.logistic, 1,0)) %>%
  summarize(misclass.rate = mean(misclassify))
  return(misclass_lr$misclass.rate)
}
```

#### Problem 8
```{r}
tree_Confusion_Matrix <- confusion_matrix(data2, data2$HardDrugs, mod_tree2)
tree_accuracy <- misclass(tree_Confusion_Matrix)
tree_Confusion_Matrix

paste("Misclassification Error Rate: ", tree_accuracy)
```

